Demography and linked selection in wild and domesticated maize

### 2015-08-14 manuscript.tex
first section of results

\section{Results}
\subsection{Patterns of diversity differ between genic and
  non-genic regions of the genome}
Previous research of the maize domestication process has relied upon
observations drawn from genic DNA \cite{wright2005,wang1999,eyre1998}. Our data were generated through whole genome sequencing,
which eliminated this constraint. Importantly, we observed substantial
differences in patterns of diversity between genic and non-genic regions
of the genome for both maize and teosinte. For maize, mean pairwise
diversity ($\pi$) within genes was significantly lower than at
positions at least 5 kb away from genes (0.00668 within, 0.00691 away, p$<$2e-44). The same
pattern of significantly greater diversity outside of genes was observed in
teosinte, but to a larger extent. For teosinte we observed $\pi$
within genes to be 0.0088 and at least 5 kb from genes it was 0.0115
(p$\approx$0). These observations suggest that genes are not evolving
neutrally in maize or teosinte. Instead, some form of selection is likely
reducing diversity within genes. Additionally, these observations
suggest that demographic inference, which relies on the assumption of
neutrally evolving DNA, will be more accurate if it is based on
observations from non-genic genetic material.

\subsection{Hard sweeps do not shape maize (or teosinte) diversity}
A mutation that is immediately beneficial and positively selected leaves a classical hard
sweep signature in the genome, whereby surrounding genetic diversity
is reduced as the haplotype where the mutation first
arose increases in frequency to fixation. The prevalence of such hard
sweeps was evaluated by comparing diversity
surrounding non-synonymous (specifically missense) and synonymous substitutions in maize and
teosinte since divergence from \emph{Tripsacum}, roughly 1-1.2 million years before
present \cite{ross2009}. For both taxa, no difference in diversity
surrounding these classes of substitutions was identified (Figure
\ref{fig:hardSweeps}). This observation suggests that hard sweeps are not a primary form of
selection for either maize or teosinte.

To ensure that this analysis is resilient to the intricacies of our sample, we investigated diversity surrounding the maize gene \emph{tga1}, one of the few known instances of a hard sweep corresponding to a causal missense substitution in maize \cite{wang2015}. In the case of this known sweep, we observed a reduction in diversity surrounding the causative substitution to levels much lower than seen for synonymous substitutions (Figure \ref{sFig:tga1}), confirming that this method should be informative if hard sweeps were the prevailing mode of selection in maize. Still, to ensure that the approach is robust to potential shortcomings such as limited power in the face of background selection \cite{enard2014}, we sought to determine if patterns of diversity surrounding synonymous and non-synonymous substitutions differ when we restrict our analysis to the most- or least-conserved genes in maize. When we re-analyzed subsets of genes with the 10\% highest and 10\% lowest genomic evolutionary rate profile (GERP) score, putatively corresponding to genes undergoing the strongest and weakest purifying selection \cite{davydov2010, cooper2005, rodgers2015}, we again saw no significant difference in diversity surrounding synonymous and non-synonymous sites (Figure \ref{sFig:consUncons}).

\begin{figure*}[!htb]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvM_Folded2_Significance_Aug}
\hspace{0.05\textwidth} \includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvT_Folded2_Significance_Aug}
\caption{Pairwise diversity surrounding synonymous and non-synonymous
  substitutions in maize and teosinte. {\bf A:} Maize diversity; {\bf B:} Teosinte diversity. Diversity axes show both absolute values and values relative to mean nucleotide diversity in windows $\geq 0.01 cM$ from a substitution. \jri{is this correct or did i confuse things?}
 Lines depict a loess curve (span of 0.01)  and shading represents bootstrap-based 95\% confidence intervals. Inset plots depict a larger range on the x-axis. \label{fig:hardSweeps}}
\end{figure*}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=.5\textwidth]{FigsAndFiles/SFS_and_Tajima.png}
\caption{The folded SFS and Tajimas D were calculated both inside and outside of genes for maize and teosinte. In both taxa, a dearth of rare alleles was observed within genes relative to outside of genes, which led to higher values of Tajima's D outside of genes than inside.{\bf A:} Folded SFS for maize; {\bf B:} Folded SFS for teosinte; {\bf C:} Maize Tajima's D; {\bf D:} Teosinte Tajima's D.  \label{fig:sfs} }
\end{figure}

@@

\section{Post-domestication expansion is apparent in patterns of singleton diversity}
Motivated by the rapid post-domestication expansion in maize that is evident from our demographic analysis, we conducted an analysis of genome-wide patterns of singleton diversity. Singleton diversity measures the abundance of alleles that are only observed in one individual among the sample. As a class, singleton alleles depict the most recent patterns of evolution, but
they also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Since our sample size
for maize (n=23) was larger than teosinte (n=13), maize singletons were
analyzed directly as well as down-sampled to reflect the sample size of
teosinte. For singletons, our definition of
neutral diversity was defined as diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged compared to what we saw with respect to pairwise diversity. Maize singleton diversity was at least as reduced as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{fig:purify}), implying that in the
recent past maize has been more influenced by purifying selection than has
teosinte.

@@

\section{Discussion}
\subsection{Little positive selection on new genic mutations}
Our findings indicate that hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize. More specifically, we observe that positive selection on new genic mutations in maize is not common compared to other drivers of diversity. This finding does not exclude the possibility that hard sweeps have taken place at non-genic sites, such as in those in enhancer regions. One known example of positive selection on a non-genic mutation involves the \emph{tb1} locus of maize, which is one of the best characterized examples of positive selection on a ``domestication gene'' in any crop \cite{clark2006}. However, the maize \emph{tb1} allele was already present in teosinte before domestication \cite{studer2011}, so this this example is in agreement with our finding that hard sweeps are rare. The \emph{gt1} locus is another well-characterized case of positive selection operating on standing variation at an enhancer region \cite{wills2013}. Instances of selection on standing variation such as these, often called soft sweeps, may be a major contributor to maize patterns of diversity \cite{beissinger2014}, and this could be part of the explanation as to why hard sweeps appear to be so rare, despite obvious morphological differences between maize and teosinte.

@@

\subsection{Demography of domestication}
Since we observed that classical hard sweeps do not explain patterns of maize diversity, it is perhaps unsurprising that we instead found evidence of demographic history contributing substantially to genetic diversity. Our estimate of the demographic history of maize, which is the first estimate of maize demographic history to utilize whole-genome-sequence (WGS) data, allowed us to disentangle the confounding parameters of bottleneck strength and bottleneck intensity. This confounding limited previous domestication estimates \cite{wright2005}. By utilizing the two-dimensional site frequency spectrum (SFS) between maize and teosinte for inference \cite{gutenkunst2009} we estimate that the maize effective population size during the bottleneck was approximately 5\% that of teosinte. A bottleneck of this magnitude may have enabled moderately deleterious alleles to reach intermediate frequency in maize, while also increasing the probability that strongly deleterious alleles segregating at low frequency in teosinte were purged from the maize gene pool, due to elevated drift as is predicted during a domestication bottleneck \cite{tajima1989}.

>>>>

\section{Results}
\subsection{Patterns of diversity differ between genic and
  non-genic regions of the genome}
Our reanalysis of the maize HapMap 2 data \cite{chia2012} is consistent with earlier findings that patterns of diversity differ between genic and non-genic regions of the genome in both maize and teosinte \cite{hufford2012} (Figure \ref{fig:diversity}). 
In maize, mean pairwise diversity ($\pi$) within genes was significantly lower than at positions at least 5 kb away from genes (0.00668 vs 0.00691, $p<2\times 10^{-44}$). 
Diversity differences in teosinte are even more pronounced (0.0088 vs. 0.0115, $p\approx 0$). 
Differences were also apparent in the site frequency spectrum, with mean Tajima's D positive in genic regions in both maize (0.4) and teosinte (0.013) but negative outside of genes (-0.087 in maize and -0.25 in teosinte, $p<X$ for both comparisons).
These observations suggest that diversity in genes is not evolving neutrally, but instead reduced by the impacts of selection on linked sites.

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{FigsAndFiles/SFS_and_Tajima.png}
\caption{The folded SFS and Tajimas D were calculated both inside and outside of genes for maize and teosinte. In both taxa, a dearth of rare alleles was observed within genes relative to outside of genes, which led to higher values of Tajima's D outside of genes than inside.{\bf A:} Folded SFS for maize; {\bf B:} Folded SFS for teosinte; {\bf C:} Maize Tajima's D; {\bf D:} Teosinte Tajima's D.  \label{fig:diversity} \jri{skip the SFS and make this a panel plot of pi and TajD in maize and teosinte} } 
\end{figure}

\subsection{Hard sweeps do not explain diversity differences}
Selection acting to increase the frequency of a new beneficial mutation will leave a signature of reduced diversity at surrounding linked sites \citex.
To evaluate whether patterns of such ``hard sweeps'' could explain observed differences in diversity between genic and non-genic regions of the genome, we compared diversity around missense and synonymous substitutions between \emph{Tripsacum} and either maize or teosinte (Figure
\ref{fig:hardSweeps}).
If a proportion of missense mutations have been fixed due to hard sweeps, diversity around these substitutions should be lower than around synonymous substitutions. 
We observe this pattern around the causative amino acid substitution in the the domestication locus \emph{tga1} (Figure \ref{sFig:tga1}), likely the result of a hard sweep during domestication \cite{wang2005origin, wang2015}.
Genome-wide, however, we observe no differences in diversity between synonymous and missense substitutions in either maize or teosinte, suggesting that hard sweeps do not play a major role in patterning diversity in genes.
Previous analyses in humans have suggested that this approach is limited by fact that a higher proportion of nonsynonymous substitutions will be found in genes under weak purifying selection and thus higher genetic diversity \cite{enard2014}. 
To address this concern we took advantage of a measure of evolutionary constraint using genomic evolutionary rate profile (GERP) scores \cite{davydov2010} calculated across the maize genome \cite{rodgers2015}. 
We re-analyzed substitutions in subsets of genes with the highest and lowest 10\% quantile of mean GERP score, putatively representing genes under the strongest and weakest purifying selection  (Figure \ref{sFig:consUncons}). 
As expected, we see higher overall diversity around substitutions in genes under weak purifying selection, but we still see no difference between synonymous and missense substitutions in either subset of the data.

%To ensure that this analysis is resilient to the intricacies of our sample, we investigated diversity surrounding the maize gene \emph{tga1}, one of the few known instances of a hard sweep corresponding to a causal missense substitution in maize \cite{wang2015}. In the case of this known sweep, we observed a reduction in diversity surrounding the causative substitution to levels much lower than seen for synonymous substitutions (Figure \ref{sFig:tga1}), confirming that this method should be informative if hard sweeps were the prevailing mode of selection in maize. 
%Still, to ensure that the approach is robust to potential shortcomings such as limited power in the face of background selection \cite{enard2014}, we sought to determine if patterns of diversity surrounding synonymous and non-synonymous substitutions differ when we restrict our analysis to the most- or least-conserved genes in maize. When we re-analyzed subsets of genes with the 10\% highest and 10\% lowest genomic evolutionary rate profile (GERP) score, putatively corresponding to genes undergoing the strongest and weakest purifying selection \cite{davydov2010, cooper2005, rodgers2015}, we again saw no significant difference in diversity surrounding synonymous and non-synonymous sites (Figure \ref{sFig:consUncons}).

\begin{figure*}[!htb]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvM_Folded2_Significance_Aug}
\hspace{0.05\textwidth} \includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvT_Folded2_Significance_Aug}
\caption{Pairwise diversity surrounding synonymous and non-synonymous
  substitutions in {\bf A} maize and {\bf B} teosinte. Axes show both absolute diversity values and values relative to mean nucleotide diversity in windows $\geq 0.01 cM$ from a substitution. \jri{is this correct or did i confuse things?}
 Lines depict a loess curve (span of 0.01)  and shading represents bootstrap-based 95\% confidence intervals. Inset plots depict a larger range on the x-axis. \label{fig:hardSweeps}}
\end{figure*}

@@

@@

\subsection{Post-domestication expansion is apparent in patterns of singleton diversity}
Motivated by the rapid post-domestication expansion in maize that is evident from our demographic analysis, we conducted an analysis of genome-wide patterns of singleton diversity. Singleton diversity measures the abundance of alleles that are only observed in one individual among the sample. As a class, singleton alleles depict the most recent patterns of evolution, but
they also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Since our sample size
for maize (n=23) was larger than teosinte (n=13), maize singletons were
analyzed directly as well as down-sampled to reflect the sample size of
teosinte. For singletons, our definition of
neutral diversity was defined as diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged compared to what we saw with respect to pairwise diversity. Maize singleton diversity was at least as reduced as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{fig:purify}), implying that in the
recent past maize has been more influenced by purifying selection than has
teosinte.

@@

\section{Discussion}
\subsection{Little positive selection on new genic mutations}
\jri{mention divergence time from \emph{Tripsacum}, roughly 1-1.2 million years before present \cite{ross2009}.}Our findings indicate that hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize. More specifically, we observe that positive selection on new genic mutations in maize is not common compared to other drivers of diversity. This finding does not exclude the possibility that hard sweeps have taken place at non-genic sites, such as in those in enhancer regions. One known example of positive selection on a non-genic mutation involves the \emph{tb1} locus of maize, which is one of the best characterized examples of positive selection on a ``domestication gene'' in any crop \cite{clark2006}. However, the maize \emph{tb1} allele was already present in teosinte before domestication \cite{studer2011}, so this this example is in agreement with our finding that hard sweeps are rare. The \emph{gt1} locus is another well-characterized case of positive selection operating on standing variation at an enhancer region \cite{wills2013}. Instances of selection on standing variation such as these, often called soft sweeps, may be a major contributor to maize patterns of diversity \cite{beissinger2014}, and this could be part of the explanation as to why hard sweeps appear to be so rare, despite obvious morphological differences between maize and teosinte.

@@

\subsection{Demography of domestication}
\jri{add note that demography inference will require regions far from genes, compare to \cite{wright2005,wang1999,eyre1998}. cite some theory or empirical paper showing this to be the case in humans/drosophila}
Since we observed that classical hard sweeps do not explain patterns of maize diversity, it is perhaps unsurprising that we instead found evidence of demographic history contributing substantially to genetic diversity. Our estimate of the demographic history of maize, which is the first estimate of maize demographic history to utilize whole-genome-sequence (WGS) data, allowed us to disentangle the confounding parameters of bottleneck strength and bottleneck intensity. This confounding limited previous domestication estimates \cite{wright2005}. By utilizing the two-dimensional site frequency spectrum (SFS) between maize and teosinte for inference \cite{gutenkunst2009} we estimate that the maize effective population size during the bottleneck was approximately 5\% that of teosinte. A bottleneck of this magnitude may have enabled moderately deleterious alleles to reach intermediate frequency in maize, while also increasing the probability that strongly deleterious alleles segregating at low frequency in teosinte were purged from the maize gene pool, due to elevated drift as is predicted during a domestication bottleneck \cite{tajima1989}.
